---
title: "GeneTrajectory example (Liver zonation & circadian effects)"
author: "Rihao Qu"
date: "9/20/2023"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_chunk$set(fig.width=5, fig.height=4.5) 
```

### Preparation
```{r, warning = FALSE, message=FALSE}

##### Load required R libraries
library(Seurat)
require(scales)
require(ggplot2)
require(viridis)
require(dplyr)
require(GeneTrajectory)
require(Matrix)
require(plot3D)
```


### Loading example data

```{r, warning = FALSE, fig.width=4.5, fig.height=4.5}
##### Import the example SMC data and take a preliminary visualization
data_S <- readRDS("/banach1/rq25/GeneTrajectory_data/data_S_liver_subset.rds")
data_S$ZT <- "6h"
data_S$ZT[which(data_S$orig.ident == "ZT00B")] <- "0h"
data_S$ZT[which(data_S$orig.ident == "ZT12B")] <- "12h"
data_S$ZT <- factor(data_S$ZT, levels = c("0h", "6h", "12h"))
DimPlot(data_S, group.by = "ZT", shuffle = T)
DimPlot(data_S, group.by = "layer", shuffle = T)
```

################################################
### Gene-gene distance computation
################################################
##### Select genes
```{r, warning = FALSE}
assay <- "RNA"
DefaultAssay(data_S) <- assay
data_S <- FindVariableFeatures(data_S, nfeatures = 2000)
all_genes <- data_S@assays[[assay]]@var.features
assay <- "alra"
expr_percent <- apply(as.matrix(data_S[[assay]]@data[all_genes, ]) > 0, 1, sum)/ncol(data_S)
genes <- all_genes[which(expr_percent > 0.05 & expr_percent < 0.75)]
length(genes)
```

##### Prepare the input for gene-gene Wasserstein distance computation
```{r, warning = FALSE}
cell.graph.dist <- GetGraphDistance(data_S, K = 10, dims = 1:2, reduction = "umap")
cg_output <- CoarseGrain(data_S, cell.graph.dist, genes, N = 1000, dims = 1:2, reduction = "umap", assay = "RNA")

#####Save the files for computation in Python
dir.path <- "/banach1/rq25/tmp/liver_zonation_example/"
dir.create(dir.path, recursive = T)
setwd(dir.path)
write.table(cg_output[["features"]], paste0(dir.path, "gene_names.csv"), row.names = F, col.names = F, sep = ",")
write.table(cg_output[["graph.dist"]], paste0(dir.path, "ot_cost.csv"), row.names = F, col.names = F, sep = ",")
Matrix::writeMM(Matrix(cg_output[["gene.expression"]], sparse = T), paste0(dir.path, "gene_expression.mtx"))
```

################################################
##### Wasserstein distance computation in Python
################################################
```{r, warning = FALSE, fig.width=5, fig.height=4.5}
# Make sure to install the latest version of POT module (python), using the following:
# pip install -U https://github.com/PythonOT/POT/archive/master.zip
# Run the following command to get the gene-gene Wasserstein distance matrix

#system(sprintf("nohup /data/rihao/anaconda3/bin/python /data/rihao/GeneTrajectory/GeneTrajectory/python/gene_distance_cal_parallel.py %s &", dir.path))
```

################################################
### Gene trajectory inference and visualization
################################################
```{r, warning = FALSE, fig.width=4.5, fig.height=4}
dir.path <- "/banach1/rq25/UMAP_GD-k5-dim2_CG-N500-dim2/"
gene.dist.mat <- LoadGeneDistMat(dir.path, file_name = "emd.csv")[genes, genes]
dim(gene.dist.mat)

gene_embedding <- GetGeneEmbedding(gene.dist.mat, K = 5)$diffu.emb

gene_trajectory <- ExtractGeneTrajectory(gene_embedding, gene.dist.mat, N = 2, t.list = c(3,5), K = 5)
table(gene_trajectory$selected)

par(mar = c(1.5,1.5,1.5,1.5))
scatter3D(gene_embedding[,1],
          gene_embedding[,2],
          gene_embedding[,3],
          bty = "b2", colvar = as.integer(as.factor(gene_trajectory$selected))-1,
          main = "trajectory", pch = 19, cex = 1, theta = 0, phi = 45,
          col = ramp.col(c(hue_pal()(2))))
scatter3D(gene_embedding[,1],
          gene_embedding[,2],
          gene_embedding[,4],
          bty = "b2", colvar = as.integer(as.factor(gene_trajectory$selected))-1,
          main = "trajectory", pch = 19, cex = 1, theta = 0, phi = 45,
          col = ramp.col(c(hue_pal()(2))))
scatter3D(gene_embedding[,2],
          gene_embedding[,3],
          gene_embedding[,4],
          bty = "b2", colvar = as.integer(as.factor(gene_trajectory$selected))-1,
          main = "trajectory", pch = 19, cex = 1, theta = 0, phi = 45,
          col = ramp.col(c(hue_pal()(2))))

gene_list <- list()
for (i in 1:2){
  #message(paste0("Trajectory-", i))
  gene_trajectory_sub <- gene_trajectory[which(gene_trajectory$selected == paste0("Trajectory-", i)),]
  genes <- rownames(gene_trajectory_sub)[order(gene_trajectory_sub[, paste0("Pseudoorder", i)])]
  #message(paste(genes, collapse = ", "))
  gene_list[[i]] <- genes
}

```


### Visualize gene bin plots
```{r, warning = FALSE, fig.width=12, fig.height=3}

data_S <- AddGeneBinScore(data_S, gene_trajectory, N.bin = 5, trajectories = 1:2, assay = "alra", reverse = c(T, T))
FeaturePlot(data_S, pt.size = 0.05, features = paste0("Trajectory",1,"_genes", 1:5), ncol = 5, order = T) &
  scale_color_gradientn(colors = rev(brewer_pal(palette = "RdYlBu")(10))) & NoLegend() & NoAxes()
FeaturePlot(data_S, pt.size = 0.05, features = paste0("Trajectory",2,"_genes", 1:5), ncol = 5, order = T) &
  scale_color_gradientn(colors = rev(brewer_pal(palette = "RdYlBu")(10))) & NoLegend() & NoAxes()

```

### Examining the association between each gene trajectory and liver zonation / circadian effects
```{r, warning = FALSE, fig.width=12, fig.height=3}
layer_matrix <- matrix(0, nrow = 8, ncol = ncol(data_S))
for (i in 1:8){
  layer_matrix[i,which(data_S$layer == as.character(i))] <- 1
}
layer_matrix <- layer_matrix/apply(layer_matrix, 1, sum)
rownames(layer_matrix) <- paste0("Layer", 1:8)

df.plot <- data.frame(layer = rep(paste0("Layer", 1:8), 5),
                      bin = rep(1:5, each = 8),
                      score = 0)

mdata <- data_S@meta.data[, paste0("Trajectory1_genes",1:5)]
mdata_by_layer <- layer_matrix %*% as.matrix(mdata)
df.plot$score <- as.vector(mdata_by_layer)
ggplot(df.plot, aes(x = layer, y = score, fill = layer)) + facet_wrap(~bin, ncol = 5) + geom_bar(stat="identity") + ggtitle("Trajectory1 - zonation") & RotatedAxis()

mdata <- data_S@meta.data[, paste0("Trajectory2_genes",1:5)]
mdata_by_layer <- layer_matrix %*% as.matrix(mdata)
df.plot$score <- as.vector(mdata_by_layer)
ggplot(df.plot, aes(x = layer, y = score, fill = layer)) + facet_wrap(~bin, ncol = 5) + geom_bar(stat="identity") + ggtitle("Trajectory2 - zonation") & RotatedAxis()
```

```{r, warning = FALSE, fig.width=5, fig.height=2.5}
circadian_matrix <- matrix(0, nrow = 3, ncol = ncol(data_S))
for (i in 1:3){
  circadian_matrix[i,which(data_S$ZT == c("0h","6h","12h")[i])] <- 1
}
circadian_matrix <- circadian_matrix/apply(circadian_matrix, 1, sum)
rownames(circadian_matrix) <- c("0h","6h","12h")

df.plot <- data.frame(ZT = rep(c("0h","6h","12h"), 5),
                      bin = rep(1:5, each = 3),
                      score = 0)
df.plot$ZT <- factor(df.plot$ZT , levels = c("0h","6h","12h"))

mdata <- data_S@meta.data[, paste0("Trajectory1_genes",1:5)]
mdata_by_circadian <- circadian_matrix %*% as.matrix(mdata)
df.plot$score <- as.vector(mdata_by_circadian)
ggplot(df.plot, aes(x = ZT, y = score, fill = ZT)) + facet_wrap(~bin, ncol = 5) + geom_bar(stat="identity") + ggtitle("Trajectory1 - circadian") & RotatedAxis()

mdata <- data_S@meta.data[, paste0("Trajectory2_genes",1:5)]
mdata_by_circadian <- circadian_matrix %*% as.matrix(mdata)
df.plot$score <- as.vector(mdata_by_circadian)
#ggplot(df.plot, aes(x = bin, y = score, fill = ZT)) + geom_bar(stat="identity", position = "dodge") + ggtitle("Trajectory2 - circadian") & RotatedAxis()
ggplot(df.plot, aes(x = ZT, y = score, fill = ZT)) + facet_wrap(~bin, ncol = 5) + geom_bar(stat="identity") + ggtitle("Trajectory2 - circadian") & RotatedAxis()

```

